Choice of plant size proxy

### 2021-04-19 doc/choose_size_var.Rmd
pare appendix down to essentials

---
title: "Untitled Draft"
author: "Report Author"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, paged.print = FALSE)
## target knits Rmds in their own session, so load libraries here.
source(here::here("packages.R"))
withr::with_dir(here::here(), {
  targets::tar_load(model_data)
})
library(performance)
```

\#TODO: The last part with the transplant experiment data is probably enough.
Probably delete the GAM comparisons and make the last bit work with `targets` by adding that data file as a `target`.

@@

The goal is to determine whether plant height, number of shoots, or their product (shoots \* height) is a good proxy for plant biomass.
First, using the 10 year demographic dataset, we can see how vital rate models fit using each of these as a predictor perform.

## Set up

Create columns for log-transformed size variables.

```{r}
df <- 
  model_data %>% 
  mutate(log_ht_prev = log(ht_prev),
         log_ht = log(ht),
         log_shts_prev = log(shts_prev),
         log_shts = log(shts))
```

Set up parallel computation

```{r}
ncores <- detectCores()
cl <- makeForkCluster(ncores - 2)
```

### Build formulas

```{r}
s_size <- surv ~ 
  s(log_size_prev, bs = "cr") +
  s(plot, bs = "re") + #random intercept
  s(spei_history, L,
    bs = "cb",
    k = c(3, 35),  #bimodal response (at most) to drought, one knot per month lag.
    xt = list(bs = "cr"))

s_ht <- surv ~ 
  s(log_ht_prev, bs = "cr") +
  s(plot, bs = "re") + #random intercept
  s(spei_history, L,
    bs = "cb",
    k = c(3, 35),  #bimodal response (at most) to drought, one knot per month lag.
    xt = list(bs = "cr"))

s_shts <- surv ~ 
  s(log_shts_prev, bs = "cr") +
  s(plot, bs = "re") + #random intercept
  s(spei_history, L,
    bs = "cb",
    k = c(3, 35),  #bimodal response (at most) to drought, one knot per month lag.
    xt = list(bs = "cr"))
```

## Survival

```{r}
m_size <- 
  bam(s_size,
      family = binomial,
      data = df,
      method = "REML",
      select = TRUE,
      cluster = cl)

m_ht <- 
  bam(s_ht,
      family = binomial,
      data = df,
      method = "REML",
      select = TRUE,
      cluster = cl)

m_shts <- 
  bam(s_shts,
      family = binomial,
      data = df,
      method = "REML",
      select = TRUE,
      cluster = cl)
```

### Compare performance

```{r}
compare_performance(m_size, m_ht, m_shts, metrics = c("AIC", "R2", "RMSE"), rank = TRUE)
```

log(size) has lowest AIC, similar R2 to height.
Number of shoots is worst fit to survival.

## Flowering

```{r}
f_ht <- update(s_ht, flwr ~ .)
f_size <- update(s_size, flwr ~ .)
f_shts <- update(s_shts, flwr ~ .)
```

Use only live plants for flowering probability.

```{r}
df2 <- df %>% filter(surv == 1)
```

```{r}
m_size <- 
  bam(f_size,
      family = binomial,
      data = df2,
      method = "REML",
      select = TRUE,
      cluster = cl)

m_ht <- 
  bam(f_ht,
      family = binomial,
      data = df2,
      method = "REML",
      select = TRUE,
      cluster = cl)

m_shts <- 
  bam(f_shts,
      family = binomial,
      data = df2,
      method = "REML",
      select = TRUE,
      cluster = cl)
```

### Compare performance

```{r}
compare_performance(m_size, m_ht, m_shts, metrics = c("AIC", "R2", "RMSE"), rank = TRUE)
```

## Data approach

Which is a better predictor of leaf area, stem number, height, or size (stem number\* height)?

Using data from transplant experiment (CITATION)

```{r}
library(readxl)
library(janitor)
la <- read_excel(here("data", "HA-la-stems-ht.xlsx")) %>% clean_names()
la <- 
  la %>%
  rename(leaf_area = leaf_area_feb_99,
         leaves = leaves_feb_99,
         ht = height_feb_99,
         shts = stems_feb_99) %>% 
  mutate(size = shts * ht) %>% 
  mutate(across(c(leaf_area, leaves, ht, shts, size), list(log = log), .names = "{.fn}_{.col}")) %>% 
  mutate(log_leaves_ht = log(leaves * ht))

la$exp %>% unique()
```

```{r}
ggplot(la, aes(x = ht, y = leaf_area)) + geom_point()
ggplot(la, aes(x = shts, y = leaf_area)) + geom_point()
ggplot(la, aes(x = size, y = leaf_area)) + geom_point()
ggplot(la, aes(x = leaves, y = leaf_area)) + geom_point()
```

I can already tell size is going to be the best

```{r}
hist(la$leaf_area)
hist(la$log_leaf_area)
```

I'll use log_leaf_area as response

@@

Options we don't have (because we don't have leaf number), but might be better proxies of plant size:

```{r}
m_leaves <- lm(log_leaf_area ~ log_leaves, data = la)
m_leaves_ht <- lm(log_leaf_area ~ log_leaves_ht, data = la)
m_leaves_shts <- lm(log_leaf_area ~ I(log(leaves*shts)), data = la)
```

```{r}
AIC(m_leaves_ht)
```

```{r}
compare_performance(m_ht, m_shts, m_size, m_leaves, m_leaves_ht,m_leaves_shts, metrics = "common", rank = TRUE)
```

>>>>

---
title: "Choice of plant size proxy"
author: "Eric R. Scott"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, paged.print = FALSE)

withr::with_dir(here::here(), {
  targets::tar_load(la_data)
})

library(tidyverse)
library(performance)
```

@@

In our 10 year demographic dataset we have yearly measurements of plant height (height to tip of tallest leaf) and shoot number.
Height $\times$ shoot number is a possibility for an integrative proxy of plant size.

Using data from transplant experiments (CITATION), we show that plant size (height $\times$ shoot number) best explains variation in total plant leaf area.

```{r}
la <- 
  la_data %>% 
  mutate(size = shts * ht) %>% 
  #log transform all size variables
  mutate(across(c(leaf_area, leaves, ht, shts, size), list(log = log), .names = "{.fn}_{.col}")) %>% 
  filter(!is.na(leaf_area) & !is.na(ht) & !is.na(shts)) %>% 
  filter(shts > 0)
```

```{r}
hist(la$leaf_area)
hist(la$log_leaf_area)
#use log(leaf area) as response to better meet normality assumptions.
```

@@

```{r}
check_model(m_ht)
check_model(m_shts)
check_model(m_size)
```

```{r}
compare_performance(m_ht, m_shts, m_size, metrics = "common", rank = TRUE)
```

```{r}
plot_fit <- function(m) {
  df <- broom::augment(m, se_fit = TRUE)

  pred <- colnames(df)[2]
  resp <- colnames(df)[1]

  p <- ggplot(df, aes_string(x = pred)) +
    geom_point(aes_string(y = resp)) + 
    geom_ribbon(aes(ymin = .fitted - .se.fit, ymax = .fitted + .se.fit), alpha = 0.4) +
    geom_line(aes(y = .fitted))
  return(p)
}
```

```{r}
plot_fit(m_ht) + labs(x = "log(height)", y = "log(leaf area)")
plot_fit(m_shts) + labs(x = "log(shoot number)", y = "log(leaf area)")
plot_fit(m_size) + labs(x = "log(shoots * height)", y = "log(leaf area)")
```

Using the log of size (shoot number $\times$ height) as a predictor gives the highest $R^2$ and lowest RMSE of the models as well as the lowest AIC indicating that it results in the best explanatory power and model fit for predicting log(leaf area) as a response.
